Linking microbial communities to ecosystem functions: what we can learn from genotype-phenotype mapping in organisms

### 2019-11-12 manuscript/manuscript_rev1.Rmd
Incorporating Brendan's edits

Microorganisms mediate many important ecosystem functions, yet it remains
unclear to what extent microbial diversity or community composition is
important for determining the rates of ecosystem-scale functions. This
uncertainty limits our ability to predict and manage crucial
microbially mediated processes, such as nutrient loss and greenhouse gas
emissions. Our lack of understanding stems from the relatively large
diversity of microorganisms, the difficulty in directly identifying
functional groups, and our limited ability to manipulate microbial community
attributes. For this reason, we propose that integrating traditional
biodiversity-ecosystem function research with ideas from genotype-phenotype
mapping could provide the new perspective our discipline needs. We identify
three insights from genotype-phenotype mapping that could be useful for
microbial biodiversity-ecosystem function studies: the concept of “agnostic”
mapping, the use of more powerful ways to account for multiple comparisons,
and the incorporation of covariates into models of ecosystem function. We
illustrate the potential for these approaches to elucidate microbial
biodiversity-ecosystem function relationships by analyzing a subset of
published data measuring methane oxidation rates from incubations of tropical
soil. We assert that combining the approaches of traditional
biodiversity-ecosystem function research with ideas from genotype-phenotype
mapping will not only generate novel hypotheses about how complex microbial
communities drive ecosystem function, but also help scientists predict and
manage changes to ecosystem functions resulting from human activities.

@@

### A new perspective on microbial community-ecosystem function relationships is needed

Ecologists have long investigated the effects of changing biodiversity on
ecosystem function, documenting, for example, relationships between terrestrial
plant community richness and primary productivity [@hooper2005]. The study of
the relationship between microbial biodiversity and ecosystem functions is much
more recent and has generated mixed results [@schimel1998, @singh2010]. Despite
the fact that microbes mediate many important ecosystem functions, it remains
maddeningly unclear to what extent microbial diversity or community composition
is important for determining the rates of ecosystem-scale functions.  For
example, numerous studies have attempted to correlate microbial functional
group abundance or diversity with the rate of various ecosystem functions and
most are unsuccessful [@rocca2015, @graham2016]. This uncertainty limits our
ability to predict and manage crucial microbially mediated ecosystem functions,
such as nutrient loss and greenhouse gas emissions. As a discipline, we need a
new perspective on these questions.

Our lack of understanding stems from many sources. Microbial communities are
more diverse and contain greater numbers of individuals than plant communities and are thus more
difficult to sample comprehensively. Unlike plants, functional groups of
microbes are rarely determined through observation or direct measurement, but
rather must be indirectly inferred based on taxonomic assignment of marker
genes (such as 16S rRNA genes), by sequencing functional genes
from metagenomes, or by quantifying specific functional genes using qPCR. This adds uncertainty in our ability to estimate the size of functional groups.
The ability to experimentally manipulate microbial diversity or community
composition is much more limited for microbes. Given all of these differences
(and many others), the challenge of linking microbial communities to ecosystem
functions may be less like that of plant communities, and more analogous to the
complex task of linking genomic variation to organismal phenotypes. For this
reason, we propose that integrating traditional biodiversity-ecosystem function
research with ideas from genotype-phenotype mapping could provide the new
perspective our discipline needs.

### There is evidence that microbial diversity can matter

Perhaps our uncertainty arises because there is no general relationship between
microbial biodiversity and any particular ecosystem function.  For example, if
local selection always optimizes the available microbial biodiversity to
maximize ecosystem function, then microbial community composition should not
matter for predicting the rates of microbially mediated functions. Instead, the
rates of these functions should be determined by the underlying environmental
variation. In this case, the microbial community would simply act as a conduit
through which the abiotic environment alters ecosystem function. Attempts to
document a general relationship between microbial community attributes and
function would then fail (since the primary drivers are environmental).

However, there is evidence that this is not always the case, A limited number
of studies have manipulated the connection between environmental factors and
microbial community composition through “common garden” or reciprocal
transplant experiments, and they frequently report different rates of ecosystem
functions for different microbial communities under the same environmental
conditions [@reed2007].  This has been observed for plant decomposition
[@glassman2018, @strickland2009], plant phenology [@panke-buisse2015], soil
nitrogen cycling [@balser2005], and soil greenhouse gas emissions
[@cavigelli2000], among others. Therefore, variation in microbial community
composition can be associated with variation in ecosystem function, independent
of environmental variation.  So why has this been difficult to consistently
document in the field?

### Traditional approaches to quantifying this relationship have provided only minor improvements

Most comparative studies of microbial community function in the field focus on
one of two aspects of microbial community structure that are hypothesized to
predict ecosystem function. The first aspect is “functional” gene or transcript
abundance.  In this case, qPCR or shotgun metagenomic sequencing is used to
estimate the abundance of a gene or transcript that is a putative marker for a
microbial process (and thus a marker for the functional group that performs
this process). For example, the gene mcrA, which encodes a subunit of the
enzyme that performs the final step in methanogenesis, is commonly used as a
marker for methanogenesis and for the methanogen functional group. Other
examples include pmoA and methanotrophy, nifH and nitrification, and nosZ and
denitrification. The abundance of these markers is often hypothesized to be
predictive of the rate of their associated processes (for example, the
abundance of mcrA is often hypothesized to be related to the rate of
methanogenesis). There are examples where this relationship is present
[@freitag2009, @freitag2010, @schnyder2018].  However, a review of such studies
found that the abundance of a functional gene or transcript is rarely
correlated with the rate of the corresponding process and depends on the
function of interest, with most effects either negative or not significant
[@rocca2015].

The second aspect of microbial community structure hypothesized to predict
ecosystem function is taxonomic or functional diversity. Diversity is either
estimated from sequence variants of a barcode gene such as the 16S rRNA gene or
manipulated through some proxy of diversity such as sequential dilution or
varying filter sizes. Studies that experimentally manipulate diversity
generally find a relationship between the applied diversity treatment and
ecosystem functions, including for methanotrophy [@schnyder2018], phosphorus
leaching [@wagg2014], greenhouse gas emissions [@wagg2014], decomposition
[@maron2018], and nitrogen cycling [@philippot2013]. However, in many cases,
richness is confounded with other factors such as the presence or absence of
major phylogenetic groups (nematodes, fungi) or abundance of microbial cells.
In addition, assembled microbial communities pose the same challenges as
macroorganismal diversity experiments, for example the highest diversity
treatment will more likely contain the most productive taxon. A review of
microbial diversity studies shows that microbial community taxonomic and
functional diversity add little explanatory power to models of ecosystem
function [@graham2016]. Overall, functional gene abundance and community
diversity improve models of ecosystem function less than one third of the time
and increase variance explained by an average of only 8 percentage points
[@graham2016].

### Genotype-phenotype mapping as a source of inspiration

Given the relative lack of success to date, it is time to rethink how we
approach the challenge of mapping microbial community structure to ecosystem
function.  Microbial ecologists are not the only biologists who are attempting
to determine the relationship between a complex set of highly-variable data and
an aggregate function.  This kind of “many-to-one” mapping is analogous to the
challenge of identifying the genetic basis of complex traits in organismal
populations. In such “genotype-phenotype” mapping studies, a population
exhibits variation in a phenotype (e.g., height or disease state) as well as
variation in potentially thousands of single nucleotide polymorphisms (SNPs).
To identify the genetic basis for a trait, investigators sample from this
population and correlate phenotype with genotype. In most cases, there are many
more loci than individuals and we do not know whether the SNPs are causally
linked or are simply in linkage disequilibrium with a causal mutation.

There are a number of parallels between this challenge faced by organismal
biologists and that facing microbial community ecologists.  They are both
“many-to-one” mapping challenges, which involve large numbers of statistical
comparisons.  Both are attempting to identify causal relationships that are
potentially confounded by very complex patterns of covariation.  There is often
no strong expectation about which entities (i.e., which genomic regions or which
microbial taxa) are most likely to be causally-related to phenotype or
function, and thus “agnostic” approaches are needed.  Both are sensitive to
exactly how the “mapping” question is asked, and ultimately require
manipulation (of genes or taxa) to establish causation.  We describe each of
these parallels below in more detail, and provide an example of how these ideas
could be applied to microbial data.

@@

The traditional approach to microorganismal biodiversity-ecosystem function
research is to measure or manipulate the diversity of a taxonomic group (e.g.,
plants) and look for an association with the function performed by that group
(e.g., primary productivity).  In the broadest sense, we can think of plants as
a “functional group,” i.e., a group of taxa united by their ability to perform
a particular ecosystem function. Ecologists may further divide a functional
group (e.g., plants) into smaller functional groupings (e.g., forbs), defined
by the details of how they perform their particular ecosystem function.

However, for microbes our knowledge of functional groups is much more limited.
From a very limited number of cultured isolates we have a provisional
understanding of which microbes might be involved in some ecosystem functions.
And by sequencing the genomes of these isolates, we have identified genetic
markers for certain functions. But most microbial taxa remain uncultured and we
do not know the function of most microbial taxa detected in environmental
samples [@hug2016, @steen2019]. In addition, there have been recent discoveries
of functions in unexpected taxonomic groups, for example methanogenesis by
fungi and cyanobacteria, a function previously considered restricted to the
Archaea [@bizic-ionescu2019, @lenhart2012]. Because of this, it would be
prudent to look more agnostically at microbial communities to identify taxa or
groups of taxa that are important for predicting the rate of ecosystem
functions rather than assuming that the genetic markers we have provisionally
identified for a given function represent the most likely taxa involved.  This
agnostic approach is analogous to the approach of many genotype-phenotype
mapping studies (e.g.,  genome-wide association studies or GWAS), which often
look for associations between a phenotype and loci anywhere in a genome.

### Dealing with multiple comparisons and covariation

Microbial biodiversity-ecosystem function studies and genotype-phenotype
mapping studies are both “many-to-one” mapping challenges, which involve large
numbers of statistical tests.  Hundreds to thousands of tests are routinely
made per study, greatly inflating the number of false positives identified
using statistical hypothesis testing approaches.  Some microbial
biodiversity-ecosystem function studies do not correct for this, while others
use approaches that may unnecessarily inflate the false negative rate, such as
the Bonferroni correction. The Bonferroni correction has been widely considered
to be too conservative, particular for exploratory studies designed to generate
hypotheses [@perneger1998, @rothman1990, @thomas1985]. Statisticians have
developed a number of less-conservative approaches to correct for multiple
comparisons by controlling the false discovery rate in order to balance the
tradeoff between Type I and Type II errors [@benjamini1995, @storey2002].
These approaches are commonly used in genotype-phenotype mapping studies.

It is widely accepted that organisms, including microorganisms, exhibit
population stratification due to geographic and environmental separation
[@martiny2006, @wright1943]. Genome-wide association studies generally account
for population structure due to shared ancestry among cases and controls when
modeling the connection between genotype and phenotype. The classic example is
the latitudinal gradient of both height and genotypic similarity in Europe,
which results in spurious associations between human height and genetic
variation [@berg2019, @novembre2008, @sohail2019]. To correct for this
covariance structure, GWAS models incorporate genotypic similarity to correct
for shared ancestry using a variety of methods, such as principal component
correction or variance component modeling [@kang2010, @price2006]. Typically,
microbial biodiversity-ecosystem function studies do not account for population
stratification (i.e., community similarity across samples), although there are
some exceptions [@meyer2019, @qin2012]. Genome-wide association studies
generally ignores the underlying environmental and spatial distance between
samples and instead uses shared ancestry as a proxy for these variables.
However, community similarity (the community analogue of shared ancestry among
organisms) is not as tightly linked to geography or environment as is shared
ancestry among organisms, and it could be very useful to account for these
separately in microbial studies, especially if one is particularly interested
in how composition alters function independent of the underlying environmental
variation.

@@

To illustrate the ideas outlined above, we reanalyzed a subset previously
published data [@meyer2019] using a modified procedure from the original
version.  A full description of the study design, samples, and data generation
can be found in that manuscript. Briefly, these data were gathered from intact
soil cores taken from diverse ecosystems of the Congo Basin in Gabon, Africa.
Cores were incubated in the laboratory under different concentrations of
methane to identify the rates of specific methane cycling pathways. For this
example, we will analyze data from just one of these pathways, high-affinity
methane oxidation (the oxidation of atmospheric concentrations of methane),
which we will refer to below as “methane oxidation”. In addition, for
simplicity we only include amplicon sequences from the DNA-inferred community
and not the RNA-inferred community, both of which are presented in the original
paper [@meyer2019].  The data we analyzed include methane oxidation
measurements, amplicon sequence variants (ASVs; @callahan2017) generated using
DADA2 and inferred from 16S rRNA gene sequences, pmoA abundance estimates (via
qPCR), latitude and longitude, and four environmental covariates (soil
moisture, bulk density, carbon, and nitrogen).

Analyses were conducted in the R statistical environment using the phyloseq
package [@mcmurdie2013, @rcoreteam2019].  The relative abundances of ASVs were
corrected using the variance stabilizing transformation from DESeq2 [@love2014,
@mcmurdie2014]. We first test typical measures of microbial structure including
functional gene abundance and community richness, which was estimated using the
breakaway package [@willis2015]. We then demonstrate significant covariation
between community structure (estimated as Bray-Curtis distance using vegan),
environmental variation (euclidean distance), and geographic distance
(euclidean distance) using Mantel tests [@bray1957, @oksanen2019].  Finally, we
present one approach to identifying taxa which are significantly associated
with function independent of the environment by fitting variance component
models using varComp to test the relationship between relative abundance of
each ASV and methane oxidation rate [@qu2013]. To illustrate how including
different covariates (environmental, geographic, and community) can result in
different conclusions about which taxa are associated with function, we fit
this model with and without random effects variance components for
environmental similarity, geographic site ID, and Bray-Curtis similarity.
Significant taxa were determined by controlling the false discovery rate at
q-value < 0.05 [@storey2002]. Figures were created using ggplot2
[@wickham2019].

Methane oxidation rate was not significantly correlated with pmoA gene
abundance or richness (Table 1, Figure 1). We tested collinearity between each
pair of distance matrices for community, environment, and geography using
Mantel tests and estimated p-values by permutation. We found a moderate and
significant correlation between community composition and environmental
variation, geography and community composition, and geography and environmental
variation (Table 2, Figure 2). Principal coordinate plots show that beta
diversity of samples separated by site ID and by ecosystem type (wetland or
upland; Figure 2). Finally, we tested the effect of the relative abundance of
each ASV on methane oxidation rate. We found different numbers of taxa
significantly associated with methane oxidation depending on which covariates
were included in the model (Table 3).  In particular, 460 taxa were identified
with no covariates whereas 6 were identified after including all covariates.
Though we cannot infer function from 16S sequences, these 6 taxa fall into
three genera and one class with cultured representatives that are not known to
consume methane [@belova2009, @chang2011, @domeignoz-horta2019, @fritz2004].
Their effect on methane oxidation rate ranges from 0.5 to 1.5 (Figure 3).  An
effect of 1.5 means that for a one unit increase in relative abundance there is
a 1.5 increase in k (the rate of exponential decrease in methane concentration
over time).

## Discussion

In this paper, we argue that the traditional approach to microbial
biodiversity-ecosystem function research often ignores the complexity of
microbial community dynamics, in particular the complex patterns of covariation
that often arise among microbial community similarity, environmental
similarity, and spatial proximity.  We propose that a better approach would be
to learn from other complex many-to-one mapping problems in biology,
particularly genotype-phenotype mapping.  These studies often explicitly
account for covariation, they correct for multiple comparisons in powerful
ways, and they frequently take an “agnostic approach” that does not assume a
particular relationship between structure and function.

We illustrate the potential for these approaches to elucidate microbial
biodiversity-ecosystem function relationships by analyzing a subset of
published data from incubations of tropical soil.  In this example, soil cores
from different ecosystems exhibited different rates of methane oxidation, but
simple correlations between these rates and the diversity or abundance of
putative methane oxidizing bacteria were not informative.  These results agree
with recent reviews of the literature that demonstrate that such simple
approaches are often not a fruitful avenue for elucidating microbial
structure-function connections.

We then asked if the relative abundance of particular taxa were related to
ecosystem function, and we identified 460 taxa whose relative abundances were
associated with rates of methane oxidation after controlling the false
discovery rate (q-value < 0.05).  These taxa could be related to ecosystem
function in multiple ways.  The most interesting possibility is that each of
these taxa is statistically related because it is causally connected to the
function. This could be direct, for example an organism that consumes methane,
or indirect, for example an organism that regulates substrates necessary for
consumers of methane. In either of these cases, the taxon could be useful as a
biomarker of function or as an organism to investigate in order to better
understand the biological drivers of variation in methane oxidation.

Alternatively, a significant association could occur for non-causal reasons.
For example, any organism that tends to be in high abundance where methane
oxidation rates are high would be correlated with methane oxidation, even if it
has no causal relationship. This could be because such an organism is favored
under the same environmental conditions that favor methane oxidizing bacteria
(or that favor methane oxidation in general).  Such covariation can drive
associations that are not causal, but the effects of such covariation can be
reduced through a number of approaches (many developed by biologists interested
in genotype-phenotype mapping).

In our example, we showed that the abundances of microbial taxa exhibit complex
patterns of covariation with each other and with environmental conditions and
spatial location. These biogeographic patterns might be even stronger if we
could sample the populations more intensively [@meyer2018]. Once we account for this
covariation, our list of taxa associated with function was reduced to 6.  These
6 taxa include taxa that are not known to directly contribute to methane
oxidation, suggesting that the drivers of methane oxidation may be indirect,
perhaps mediated by ecological interactions among taxa from multiple functional
groups.

>>>>

Microbial physiological processes are intimately involved in nutrient cycling.
However, it remains unclear to what extent microbial diversity or community
composition is important for determining the rates of ecosystem-scale
functions. There are numerous examples where microbial diversity is correlated
with the rate of an ecosystem function, but exactly how microbial communities
“map” onto ecosystem functions remains unresolved.  This uncertainty limits our
ability to predict and manage crucial microbially mediated processes such as
nutrient loss and greenhouse gas emissions. Our lack of understanding stems
from the relatively large diversity of microorganisms, the difficulty in
directly identifying functional groups, and our limited ability to manipulate
microbial community attributes. For this reason, we propose that integrating
traditional biodiversity-ecosystem function research with ideas from
genotype-phenotype mapping could provide the new perspective our discipline
needs. We identify two insights from genotype-phenotype mapping that could be
useful for microbial biodiversity-ecosystem function studies: the concept of
“agnostic” mapping and controlling for population stratification to identify
microorganisms that uniquely contribute to ecosystem function. We illustrate
the potential for these approaches to elucidate microbial
biodiversity-ecosystem function relationships by analyzing a subset of
published data measuring methane oxidation rates from tropical soils. We assert
that combining the approaches of traditional biodiversity-ecosystem function
research with ideas from genotype-phenotype mapping will not only generate
novel hypotheses about how complex microbial communities drive ecosystem
function, but also help scientists predict and manage changes to ecosystem
functions resulting from human activities.

@@

Ecology is broadly focused on understanding biodiversity and how that
biodiversity shapes the ecosystems that humans depend on. Many ecosystem
processes essential to all of life are mediated by microorganisms and therefore
understanding the relationship between microbial biodiversity and ecosystem
function is important [@schimel1998, @singh2010]. Certain ecosystem functions
are correlated with microbial diversity, indicating that we should be able to
determine what aspects of microbial biodiversity influence ecosystem function.
However, attempts to describe that mapping have borne little fruit [@rocca2015,
@graham2016]. We argue that to overcome this challenge we should look to other
successful attempts at mapping biological variation onto higher order
processes. In particular, population genetics and the process of
genotype-phenotype mapping provide a number of potentially useful insights. For
example, genotype-phenotype mapping often makes few assumptions about the
nature of the map, i.e., it is “agnostic.” In addition, population geneticists
have developed rigorous methods for reducing potentially confounding
relationships such as geographic structuring of populations. Finally, the
ultimate goal of genotype-phenotype mapping is to identify the unique
contribution of genotype to phenotype separate from other drivers of phenotypic
variation such as the environment. Inspired by this, we suggest a reframe of
the question, "is microbial biodiversity related to ecosystem function?" to, "what
is the unique contribution of the microbiome to ecosystem function independent
of the environment?"

### There is evidence that microbial biodiversity matters for ecosystem function

There is increasing evidence for microbial biodiversity-ecosystem function
relationships. For example, there are positive correlations between microbial
diversity and ecosystem multifunctionality for a variety of ecosystems and most
major lineages of microorganisms [@jing2015, @delgado-baquerizo2016,
@delgado-baquerizo2017]. Adding microbial diversity or abundance to ecosystem
models can in some cases improve model accuracy [@graham2016]. Contrived
communities that vary in richness and communities created through sequential
dilution or varying filter size to generate differences in diversity can also
exhibit differences in ecosystem function [@philippot2013, @wagg2014, @schnyder2018,
@maron2018].  Finally, reciprocal transplant and common garden experiments that
manipulate the connection between community composition and environment reveal
differences in ecosystem function for communities of distinct
origins [@cavigelli2000, @strickland2009, @glassman2018]. Given these
relationships, many investigators have now moved on to the challenge of
identifying the aspects of microbial biodiversity (e.g., specific taxa, genes,
functional groups, etc.) that influence a given ecosystem function; however,
this has proven especially challenging.

### The mapping between microbial biodiversity and ecosystem function has been elusive

Most studies that attempt to identify the aspects of microbial biodiversity
that influence a given ecosystem function focus on “functional” gene or
transcript abundance.  In this case, qPCR or shotgun metagenomic sequencing is
used to estimate the abundance of a gene or transcript that is a putative
marker for a microbial process (and thus a marker for the functional group that
performs this process). For example, the gene _mcrA_, which encodes a subunit of
the enzyme that performs the final step in methanogenesis, is commonly used as
a marker for methanogenesis and for the methanogen functional group. Other
examples include _pmoA_ and methanotrophy, _nifH_ and nitrification, and _nosZ_ and
denitrification. The abundance of these markers is often hypothesized to be
predictive of the rate of their associated processes (for example, the
abundance of _mcrA_ is often hypothesized to be related to the rate of
methanogenesis).

Some ecosystem functions in certain ecosystems can be predicted from the
abundance or transcriptional activity of genetic markers for those functions.
For example, soil methane production and consumption can under some
circumstances be predicted from the genetic markers _mcrA_ and _pmoA_
[@freitag2009, @freitag2010, @mccalley2014]. However, for most ecosystem
functions the abundance of a functional gene or transcript is rarely positively
correlated with the rate of the corresponding process [@rocca2015]. The
cases where there is a positive correlation tend to be restricted to
agricultural ecosystems and for certain functions within the nitrogen cycle
[@rocca2015]. In general, including aspects of microbial biodiversity
(e.g., functional gene abundance or diversity) improve models of ecosystem
function less than one third of the time, and increase variance explained by an
average of only 8 percentage points over environmental variables [@graham2016].

## Genotype-phenotype mapping as a source of inspiration

In the approaches described above, microbial ecologists often use microbiome
data to infer taxonomic composition, essentially creating species lists from
data such as 16S rRNA marker genes or shotgun metagenomes. Interpreting
microbiome data in this way has allowed us to use approaches from
biodiversity-ecosystem function research (which are often focused on taxonomic
or functional groups), but it has generally not been useful for creating more
detailed descriptions of the relationship between microbial biodiversity and
ecosystem function. But this approach is not the only way one could determine
the relationship between a complex set of highly-variable data and an aggregate
function.

This kind of “many-to-one” mapping is analogous to the challenge of identifying
the genetic basis of complex traits in organismal populations. In such
“genotype-phenotype” mapping studies, a population exhibits variation in a
phenotype (e.g., height or disease state) as well as variation in potentially
thousands of single nucleotide polymorphisms (SNPs). To identify the genetic
basis for a trait, investigators sample from this population and correlate
phenotype with genotype. While some phenotypes (e.g., the propensity for
diseases such as Parkinson’s) are controlled by a single genetic locus
[@kerem1989, @macdonald1992], most traits depend on a large number of genes
that control variation in that phenotype [@hindorff2009, reich2001]. In
addition, there is often no a priori expectation about which regions of the
genome control that trait so we must search for genetic markers
throughout the genome. If a marker is significantly correlated with the
phenotype of interest, this either indicates it is inside of a gene with a
direct or indirect effect on phenotype or that it is in linkage disequilibrium
(i.e., non-random association between two alleles) with a causal gene.

There are a number of parallels between this challenge faced by organismal
biologists and that facing microbial community ecologists. They both involve
large numbers of statistical comparisons. Both are attempting to identify
causal relationships that are potentially confounded by complex patterns of
covariation. There is often no strong expectations about which entities (i.e.,
which genomic regions or which microbial genes or lineages) are most likely to
be causally-related to phenotype or function, and thus “agnostic” approaches
are needed. For some ecosystem functions, it is possible that a single
microbial taxon could substantially influence its rate. For example, methane
flux from permafrost in Sweden may be controlled by a single taxon [@mccalley2014]. But
for most ecosystem functions, there could be many taxa of small effect that
contribute to the rate of ecosystem function. Finally, both ultimately require
manipulation (of genes or taxa) to establish causation.

@@

Most microbial biodiversity-ecosystem function research up to this point has
used an approach analogous to that used by plant ecologists studying
biodiversity-ecosystem function relationships. This approach is to measure or
manipulate the diversity of a taxonomic group (e.g., plants) and look for an
association with the function performed by that group (e.g., primary
productivity). We can think of plants as a “functional group,” i.e., a group of
taxa united by their ability to perform a particular ecosystem function. For
microbes, estimating functional group abundance can be much more challenging.
From a small number of cultured isolates, we have a provisional understanding
of which microbes might be involved in some ecosystem functions. By sequencing
the genomes of these isolates, we have identified genetic markers for certain
functions, which we call “functional genes.” But most microbial taxa remain
uncultured and we do not know the function of most microbial taxa detected in
environmental samples [@hug2016, @martiny2019]. In addition, there have
been recent discoveries of functions in unexpected taxonomic groups, for
example methanogenesis by fungi and cyanobacteria, a function previously
considered restricted to the Archaea [@bizic-ionescu2019, @lenhart2012].

As illustrated earlier, these functional markers are not correlated with their
corresponding ecosystem function in most ecosystems and for most processes. In
addition, they provide little explanatory power to models of ecosystem
function. Because of this, it might be prudent to look more agnostically at
microbial communities to identify taxa, groups of taxa or genes that are
important for predicting the rates of ecosystem functions rather than assuming
that the genetic markers we have provisionally identified for a given function
represent the most likely taxa or genes involved. This agnostic approach is
analogous to the approach of many genotype-phenotype mapping studies (e.g.,
genome-wide association studies), which often look for associations between a
phenotype and loci anywhere in the genome.

Beyond finding new physiologies in unexpected places, there are other reasons
for looking agnostically. In the case of microbial functions, it may be that
the organism that performs a function is not the limiting factor for the rate
of that function. For example, the rate of soil-to-atmosphere methane flux
could be limited by methanogens or methanotrophs or the balance of the two. It
could also be limited by the bacteria that produce the fermentative byproducts
that methanogens depend on. Or there could be indirect limitation by organisms
that liberate nitrogen or phosphorus into mineral forms. In other words, the
influence of microbial communities on the rate ecosystem function could
represent a complex metabolic network much like the regulation of gene
expression in organisms that partially determine their phenotype. These broader
patterns of biodiversity-ecosystem function relationships would be invisible to
any study that focuses on putative functional groups without considering the
reads that underlie those functional groups.

### Controlling for population stratification

It is widely accepted that organisms, including microorganisms, exhibit
population stratification due to geographic and environmental separation
[@wright1943, @martiny2006]. This can lead to spurious associations between
phenotypes and genetic markers that are at high frequency in isolated
sub-populations. Association studies generally control for population
stratification by accounting for shared ancestry among organisms in a
population when modeling the connection between genotype and phenotype.
Typically, microbial biodiversity-ecosystem function studies do not account for
population stratification (i.e., community similarity among ecosystems),
although there are some exceptions [@meyer2019, @qin2012, @lloyd-price2019].
Community similarity (the community analogue of shared ancestry among
organisms) is not as tightly linked to geography or environment as is shared
ancestry, and it could be very useful to account for these separately in
microbial studies, especially if one is particularly interested in how
composition alters function independent of the underlying environmental
variation.

Genome-wide association studies correct for stratification using a variety of
methods. Generally, they ignore the underlying environmental and spatial
distance between samples and instead use shared ancestry as a proxy for local
selection and assortative mating. A common approach is to perform a regression
of phenotype and shared ancestry (computed as the first one or more principal
components of the genotype matrix) and then use the residuals from this model
as the values for phenotype in a subsequent regression using the genotypes
directly [@price2006]. This principal component correction is designed to test the
effect of individual genes after removing the effect of shared ancestry among
individuals. Another approach, employed in our example, is variance component
modeling (or mixed modeling, hierarchical modeling, etc.) where genotypic
similarity is included as a covariate in the model to control for
stratification while testing the genotype-phenotype connection [@kang2010].

If we control for covariates such as community similarity, geographic distance,
or environmental similarity, it changes the nature of our question. For
example, if we test the correlation between the relative abundance of a taxon
and the rate of methane flux we are asking “is this taxon correlated with
methane flux?” If we find a significant result, that may be because variation
in the abundance of that organism directly or indirectly contributes to methane
flux. However, it might also be that that organism lives only in ecosystems
that happen to have a high rate of ecosystem function. We would not be able to
distinguish these outcomes with this simple correlation. If we then include
environmental variables or environmental similarity as a covariate, we could
start to ask, “Is the taxon uniquely associated with function independent of
the environment?” By “uniquely associated”, we mean those taxa associated with
function irrespective of environmental conditions, local community structure,
or spatial proximity. This is not only a more specific question than is usually
asked in microbial biodiversity-ecosystem function studies, but it is also a
more appropriate one, especially if one is interested in how to incorporate
microbial community data into ecosystem models. Finally, it is interesting on
its own to understand whether microbial communities are selected by the
underlying environmental conditions to produce a particular rate of ecosystem
function (the conduit) or whether there is variation in community composition
independent of the environment that explains some variation in ecosystem
function.

@@

To illustrate the ideas outlined above, we reanalyzed a subset of previously
published data from a paper that has demonstrated one successful approach
for applying genotype-phenotype mapping to microbial communities [@meyer2019]. In our
reanalysis, we do not intend to challenge the conclusions from that paper, but
instead we want to demonstrate how to perform this type of study for microbial
ecologists unfamiliar with association studies. A full description of the study
design, samples, and data generation can be found in that manuscript. Briefly,
these data were gathered from intact soil cores taken from diverse ecosystems
of the Congo Basin in Gabon, Africa. Cores were incubated in the laboratory
under different concentrations of methane to identify the rates of specific
methane cycling pathways. For this example, we will analyze data from just one
of these pathways, high-affinity methane oxidation (the oxidation of
atmospheric concentrations of methane), which we will refer to simply as
“methane oxidation”. In addition, we only include amplicon sequences from the
DNA-inferred community and not the RNA-inferred community, both of which are
presented in the original paper (@meyer2019). The data we analyzed include methane
oxidation rates, amplicon sequence variants (ASVs; @callahan2017) generated using DADA2
and inferred from 16S rRNA gene sequences, _pmoA_ abundance estimates (via qPCR),
latitude and longitude, and four environmental covariates (soil moisture, bulk
density, carbon, and nitrogen).

Analyses were conducted in the R statistical environment using the phyloseq
package [@mcmurdie2013, @rcoreteam2019]. The relative abundances of ASVs were corrected using the variance
stabilizing transformation from DESeq2 [@love2014,
@mcmurdie2014]. We first test typical measures of
microbial structure including functional gene abundance and community richness,
which was estimated using the breakaway package [@willis2015]. We then demonstrate
significant covariation between community structure (estimated as Bray-Curtis
distance using vegan), environmental variation (euclidean distance), and
geographic distance (euclidean distance) using Mantel tests [@bray1957, @oksanen2019]. Finally, we
present one approach to identifying taxa which are significantly associated
with function independent of the environment by fitting variance component
models using varComp to test the relationship between relative abundance of
each ASV and methane oxidation rate [@kang2010, @qu2013]. To illustrate how
including different
covariates (environmental, geographic, and community) can result in different
conclusions about which taxa are associated with function, we fit this model
with and without random effects variance components for environmental
similarity, geographic site ID, and Bray-Curtis similarity. Significant taxa
were determined by controlling the false discovery rate at q-value < 0.05 [@storey2002].
Figures were created using ggplot2 [@wickham2019].

### Results & Discussion

Microbial biodiversity-ecosystem function studies typically test functional
group abundance or community richness as it relates to ecosystem function. In
our case, methane oxidation rate was not significantly correlated with the pmoA
gene abundance or 16S read richness (Table 1, Figure 1). To demonstrate that
the covariance structure of the data might alter our conclusions about which
taxa regulate ecosystem function, we tested collinearity between each pair of
distance matrices for community, environment, and geography. We found a
moderate and significant correlation between community composition and
environmental variation, geography and community composition, and geography and
environmental variation (Table 2, Figure 2). To visualize this population
stratification, principal coordinate plots show that beta diversity of samples
separated by site ID and by ecosystem type (wetland or upland; Figure 2), which
indicates substantial spatial and environmental structuring of microbial
populations. This means that the presence or abundance of certain genetic
markers (taxa) will be elevated in specific ecosystems. In this case,
high-affinity methane oxidation is typically greater in upland ecosystems than
in wetland ecosystems and so any taxa differentially abundant in uplands will
be correlated with methane oxidation regardless of their involvement in that
process. It is necessary to control for this structuring to rigorously identify
associations between taxa and function.

To demonstrate this approach, we tested the effect of the relative abundance of
each ASV on methane oxidation rate while controlling for different sets of
covariates including environment, geography, and community. After controlling
the false discovery rate, 460 unique 16S reads were identified as significantly
correlated with function when no covariates were included in the model. We
found different numbers of taxa significantly associated with methane oxidation
depending on which covariates were included in the model (Table 3). Each of
these sets of taxa represent different versions of the biodiversity-ecosystem
function mapping question. For example, controlling for environmental variation
tests for taxa that are related to function independent of environmental
conditions. Similarly, controlling for geographic distance tests for taxa that
are related to function independent of geographic proximity; geographic
proximity could influence function via unmeasured environmental variation that
is spatially structured or differences in dispersal history among sites due to
physical distance. In the model that controlled for all three covariates
(community, environment, and geography), only six ASVs were significantly
correlated with methane oxidation rate (Figure 3). These taxa could be useful
indicators for methane oxidation rate across space and different ecosystems. In
addition, researchers could manipulate their abundance in order to better
understand the biological drivers of variation in methane oxidation.

Notably, these six taxa fall into three genera and one class with cultured
representatives that are not known to consume methane [@belova2009, @chang2011,
@domeignoz-horta2019, @fritz2004]. These taxa could be
related to ecosystem function in multiple ways. The most interesting
possibility is that each of these taxa is statistically related because it is
causally connected to the function. This could be direct, for example an
organism that consumes methane, or indirect, for example an organism that
regulates substrates necessary for methane oxidizers. Alternatively, a
significant association could occur for non-causal reasons. For example, any
organism that tends to be in high abundance where methane oxidation rates are
high would be correlated with methane oxidation, even if it has no causal
relationship. This could be because such an organism is favored under the same
conditions that favor methane oxidation. Such covariation can
drive associations that are not causal, but the effects of covariation would
have been reduced by controlling for covariates in our tests.

This approach is powerful because it accounts for the covariances that arise
from the fundamental ecological processes that drive community assembly and
underlie biogeographical patterns [@martiny2006, @vellend2010]. It is also
powerful because it is
focused on a much more specific version of the “biodiversity-ecosystem
function” question, i.e., it asks “which taxa are uniquely associated with
function?” By “uniquely associated,” we mean those taxa associated with
function irrespective of environmental conditions, local community structure,
or spatial proximity. This is not only a more specific question than is usually
asked in microbial biodiversity-ecosystem function studies, but it is also a
more appropriate one, especially if one is interested in how to incorporate
microbial community data into ecosystem models. These models depend on
identifying taxa that add explanatory power beyond that provided by other
factors, such as environmental conditions. Earth system models would benefit
most from microbial bioindicators that capture variation unexplained by other,
more easily measured variables (such as certain environmental conditions). The
approach we describe is an important step toward identifying such biomarkers.

## Caveats and future directions
